Deeply Integrating C11 Code Support into Isabelle/PIDE

### 2019-10-21 C11-FrontEnd/document/generated/paper.tex
synchronize with fortesse revision 5377

[label = {background},type = {scholarly_paper.technical}, args={label = {background},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''bu''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Background: PIDE and the Isabelle Document Model%

@@

[label = {frontend_arch},type = {scholarly_paper.technical}, args={label = {frontend_arch},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''bu''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]The C11 Parser Generation Process and Architecture%

@@

  recent version $j$, for any versions $i$ and $j$ satisfying $2016 \leq i \leq j \leq 2019$ (at the
  time of writing).}%

@@

[label = {arch1},type = {scholarly_paper.technical}, args={label = {arch1},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''bu''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Generating the AST%

@@

[label = {lexer},type = {scholarly_paper.technical}, args={label = {lexer},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''fred''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Constructing a Lexer for C11%

@@

[label = {parser},type = {scholarly_paper.technical}, args={label = {parser},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''fred''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Generating the Shift-Reduce Parser from the Grammar%

@@

[label = {ctests},type = {scholarly_paper.technical}, args={label = {ctests},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''fred''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Isabelle/C: Syntax Tests and Experimental Results%

@@

[label = {annotations},type = {scholarly_paper.technical}, args={label = {annotations},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''fred''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Generic Semantic Annotations for C%

@@

Obviously, the scheduling of these transformations of the underlying theory contexts is
non-trivial.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkupfigure*}%
[label = {C-sample3},type = {Isa_COL.figure}, args={label = {C-sample3},type = {Isa_COL.figure}, Isa_COL.figure.relative_width = {100}, Isa_COL.figure.src = {figures/A-C-Source3}, Isa_COL.figure.spawn_columns = {True}}]Advanced annotation programming%
\end{isamarkupfigure*}\isamarkuptrue%
%
\begin{isamarkupsubsection*}%
[label = {navigation},type = {scholarly_paper.technical}, args={label = {navigation},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''bu''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Navigation for Annotation Commands%
\end{isamarkupsubsection*}\isamarkuptrue%
%
\begin{isamarkuptext}%
In order to overcome the problem of syntactic ambiguity of annotations, we slightly
refine the syntax of semantic annotations by the concept of a navigation expression:
\begin{isar}
    $\mathcal{L}_\text{annot}$ $=$ $\varnothing$ | <navigation-expr> <annotation-command> $\mathcal{L}_\text{annot}$ 
\end{isar}

@@

[label = {annot1},type = {scholarly_paper.technical}, args={label = {annot1},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''fred''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Defining Annotation Commands%

@@

  \isa{thy{\isacharunderscore}load}, \isa{thy{\isacharunderscore}defn}, etc., for example: $K_\text{\emph{cmd}} =
  \isa{\isacommand{definition}\ {\isacharcolon}{\isacharcolon}\ thy{\isacharunderscore}defn} \isatt{|}
  \isa{\isacommand{lemma}\ {\isacharcolon}{\isacharcolon}\ thy{\isacharunderscore}goal{\isacharunderscore}stmt} \isatt{|} \dots $}

Since the type \isatt{c{\char`\_}parser} is isomorphic to \isatt{parser},
but accepting C tokens, one can use \isatt{C{\char`\_}Token.syntax{\char`\'}} to translate and carry
the default Isar commands \emph{inside} the \isa{\isacommand{C}{\isacartoucheopen}\ {\isachardot}{\isachardot}\ {\isacartoucheclose}} scope, such as \isa{\isacommand{lemma}} or
\isa{\isacommand{by}}. Using \isa{\isacommand{{\isasymsimeq}setup}}, one can even
define an annotation command \isa{\isacommand{C}} taking a C code as argument, as the ML
code of \isa{\isacommand{{\isasymsimeq}setup}} has type
$\alpha^{\text{AST}}~\isatt{{\char`\-}{\char`\>}\ env\ {\char`\-}{\char`\>}}~R_\text{\emph{cmd}}$ (which is enough for
calling \isatt{C{\char`\_}Annotation.command{\char`\'}} in the ML code). Here, whereas the type
\isatt{env} is always the same, the type $\alpha^{\text{AST}} \subseteq
\text{AST}^\text{C11}$ varies depending on {\footnotesize\texttt{<navigation-expr>}} (see \csname isaDof.ref\endcsname[type={text}]{annot2}).%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Note, however, that the user experience of the IDE changes when nesting commands too
deeply. In terms of error handling and failure treatment, there are some noteworthy implementation
differences between the outermost commands and C annotation commands. Naturally, the PIDE toplevel
has been optimized to maximize the error recovery and parallel execution. Inside a command, the
possibilities to mimic this behaviour are somewhat limited. Whereas the failure of an outermost
command must always be treated as an anomaly, in $\mathcal{L}_\text{annot}$, we are supposed to be
editing in a comment space: an error should be reported but not too much interruptive, because a
comment can contain for good reasons human narrative sentences. On the other hand, an error masking
itself shyly as warning can be overlooked easily, so a form of per-case compromise has to be found
to control the locally permissive set of errors. As a workaround useful during development and
debugging, we offer a further pragma for a global annotation, namely \isa{{\isacharasterisk}}
(in complement to the violet \isa{{\isacharat}}), that controls a switch between a
strict and a permissive error handling for nested annotation commands:
\begin{isar}
    <annotation-comment> $=$ // $[$ * $|$ @ $]^*$ $\mathcal{L}_\text{annot}$
    <annotation-comment> $=$ /* $[$ * $|$ @ $]^*$ $\mathcal{L}_\text{annot}$ */
\end{isar}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkupsubsection*}%
[label = {annot2},type = {scholarly_paper.technical}, args={label = {annot2},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''fred''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Evaluation Order%
\end{isamarkupsubsection*}\isamarkuptrue%
%
\begin{isamarkupfigure*}%
[label = {C-sample5},type = {Isa_COL.figure}, args={label = {C-sample5},type = {Isa_COL.figure}, Isa_COL.figure.relative_width = {62}, Isa_COL.figure.src = {figures/A-C-Source5}, Isa_COL.figure.spawn_columns = {True}}]Creating and nesting C annotations: the importance of evaluation order in annotations%
\end{isamarkupfigure*}\isamarkuptrue%
%
\begin{isamarkuptext}%
We will now explain why positional languages are affecting the evaluation time of
annotation commands in \csname isaDof.ref\endcsname[type={Isa_COL.figure}]{C-sample3}. This requires a little zoom on how the
parsing is actually executed.

@@

Note that the C11 grammar has enough scoping structure for the full inference of the C environment
\isatt{env} be at bottom-up time. In terms of efficiency, we use specific
\emph{static} rule wrappers having the potential of overloading default grammar rules (see \csname isaDof.ref\endcsname[type={Isa_COL.figure}]{architecture}), to assign a wrapper to be always executed as soon as a Shift-Reduce
rule node of interest is encountered. The advantage of this construction is that the wrappers are
statically compiled, which results in a very efficient reporting of C type information.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkupsection*}%
[label = {backends},type = {scholarly_paper.technical}, args={label = {backends},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''bu''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Semantic Back-Ends%
\end{isamarkupsection*}\isamarkuptrue%
%
\begin{isamarkuptext}%
In this section, we briefly present two integrations of verification back-ends for C. We
chose Clean \cite{journals/afp/TuongW19}~\footnote{stemming from recent HOL-TestGen distributions
    \url{https://www.brucker.ch/projects/hol-testgen/index.en.html}} used
for program-based test generation \cite{DBLP:conf/tap/Keller18}, and AutoCorres \cite{DBLP:conf/pldi/GreenawayLAK14}, arguably the most developed deductive verification environment
for machine-oriented C available at present.

Note that we were focusing on keeping modifications of integrated components minimal, particularly
for the case of AutoCorres. Certain functionalities like position propagation of HOL terms in
annotations, or ``automatic'' incremental declarations~\footnote{\url{https://github.com/seL4/l4v/blob/master/tools/autocorres/tests/examples/Incremental.thy}}
may require internal revisions on the back-end side. This is out of the scope of this paper.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkupsubsection*}%
[label = {clean},type = {scholarly_paper.technical}, args={label = {clean},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''bu''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]A Simple Typed Memory Model: Clean%
\end{isamarkupsubsection*}\isamarkuptrue%
%
\begin{isamarkuptext}%
Clean (pronounced as: ``C lean'' or ``Céline'' [selin]) is based on a simple,
shallow-style execution model for an imperative target language.
Importing Clean into a theory with its activated back-end proceeds as in \autoref{fig:clean}.
For a deep presentation of Clean, the reader is referred to the published AFP document \cite{journals/afp/TuongW19}.

@@

[label = {autocorres},type = {scholarly_paper.technical}, args={label = {autocorres},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''fred''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]The Case of AutoCorres%

@@

[label = {securify},type = {scholarly_paper.technical}, args={label = {securify},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''fred''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Yet another C-Back-End: Securify%

@@

In the Securify project at the Nanyang Technological University \cite{DBLP:conf/tacas/SananZHZTL17,DBLP:conf/itp/HouSTL17}, an extension of AutoCorres is in

@@

the verification framework. This ongoing project is already profiting from the overall framework,
including the integrated parser $\text{PARSE}_{\text{\emph{report}}}^{\text{C11}}$. Note, however,
that the XtratuM micro-kernel \cite{DBLP:journals/sigbed/CarrascosaCMBC14} in study is using
certain advance features of C11 not found in seL4 (including nested functions, or particular

@@

[label = {concl},type = {scholarly_paper.conclusion}, args={label = {concl},type = {scholarly_paper.conclusion}, scholarly_paper.text_section.main_author = {@{docitem ''bu''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}}]Conclusions%

@@

Isabelle/C also forms a basis for future semantically well-understood combinations of back-ends 
based on different semantic interpretations: inside Isabelle, bridge lemmas can be derived 
that describe the precise conditions under which results from one back-end can be re-interpreted
and used in another. Future tactic processes based on these bridge lemmas may open up novel ways 
for semantically safe tool combinations.%
\end{isamarkuptext}\isamarkuptrue%
%
\paragraph*{Acknowledgments.}
\begin{isamarkuptext}%
The authors warmly thank David Sanán and Yang Liu for encouraging the development and reuse of
$C^\downarrow$, initially started in the Securify project \cite{DBLP:conf/tacas/SananZHZTL17}
(\url{http://securify.sce.ntu.edu.sg/}).
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkupsection*}%
[label = {appendix},type = {scholarly_paper.technical}, args={label = {appendix},type = {scholarly_paper.technical}, scholarly_paper.text_section.main_author = {@{docitem ''fred''}}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Appendix%
\end{isamarkupsection*}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsection{Portability of Isabelle/C's Generated Files%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
To evaluate the portability of the generated files in \csname isaDof.ref\endcsname[type={Isa_COL.figure}]{architecture}, we recall that:

@@

  two ASTs was merged together while releasing version 1.18.0.} Upgrading Haskabelle to use
the new 1.18.0 was the most heavy task we performed, when integrating Haskabelle's output with
Citadelle's input. Otherwise, we have not encountered any problems in upgrading Haskabelle from
1.18.0 to more recent versions (such as 1.20.1 which was straightforward).%

@@

evaluation side (with the option \isa{quick{\isacharminus}and{\isacharminus}dirty} activated). This is the purpose of
the general \isa{\isacommand{Haskell{\isacharunderscore}file}} command implemented in Citadelle:~\footnote{Note that \isa{\isacommand{old{\isacharunderscore}datatype}} is a command, whereas
  \isa{\isakeyword{datatype{\isacharunderscore}old}} is an option provided to the command

@@

The advantage of this construction is the possibility to do immediate code value transformations
obtained from Haskabelle.

>>>>

[label = {background},type = {scholarly_paper.technical}, args={label = {background},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Background: PIDE and the Isabelle Document Model%

@@

[label = {frontend_arch},type = {scholarly_paper.technical}, args={label = {frontend_arch},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]The C11 Parser Generation Process and Architecture%

@@

  recent version $j$, for at least any versions $i$ and $j$ satisfying $2016 \leq i \leq j \leq
  2019$ (at the time of writing).}%

@@

[label = {arch1},type = {scholarly_paper.technical}, args={label = {arch1},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Generating the AST%

@@

[label = {lexer},type = {scholarly_paper.technical}, args={label = {lexer},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Constructing a Lexer for C11%

@@

[label = {parser},type = {scholarly_paper.technical}, args={label = {parser},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Generating the Shift-Reduce Parser from the Grammar%

@@

[label = {ctests},type = {scholarly_paper.technical}, args={label = {ctests},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Isabelle/C: Syntax Tests and Experimental Results%

@@

[label = {annotations},type = {scholarly_paper.technical}, args={label = {annotations},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Generic Semantic Annotations for C%

@@

Obviously, the scheduling of these transformations of the underlying theory contexts is
non-trivial.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkupfigure*}%
[label = {C-sample3},type = {Isa_COL.figure}, args={label = {C-sample3},type = {Isa_COL.figure}, Isa_COL.figure.relative_width = {100}, Isa_COL.figure.src = {figures/A-C-Source3}, Isa_COL.figure.spawn_columns = {True}}]Advanced annotation programming%
\end{isamarkupfigure*}\isamarkuptrue%
%
\begin{isamarkupsubsection*}%
[label = {navigation},type = {scholarly_paper.technical}, args={label = {navigation},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Navigation for Annotation Commands%
\end{isamarkupsubsection*}\isamarkuptrue%
%
\begin{isamarkuptext}%
In order to overcome the problem of syntactic ambiguity of annotations, we slightly
refine the syntax of semantic annotations by the concept of a navigation expression:
\begin{isar}
    $\mathcal{L}_\text{annot}$ $=$ $\varnothing$ | <navigation-expr> <annotation-command> $\mathcal{L}_\text{annot}$ 
\end{isar}

@@

[label = {annot1},type = {scholarly_paper.technical}, args={label = {annot1},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Defining Annotation Commands%

@@

  \isa{thy{\isacharunderscore}load}, \isa{thy{\isacharunderscore}defn}, etc., for example we should have defined
  $K_\text{\emph{cmd}}$ as : $K_\text{\emph{cmd}} = \isa{\isacommand{definition}\ {\isacharcolon}{\isacharcolon}\ thy{\isacharunderscore}defn} \isatt{|} \isa{\isacommand{lemma}\ {\isacharcolon}{\isacharcolon}\ thy{\isacharunderscore}goal{\isacharunderscore}stmt}
  \isatt{|} \dots $}

Since the type \isatt{c{\char`\_}parser} is isomorphic to \isatt{parser},
but accepting C tokens, one can use \isatt{C{\char`\_}Token.syntax{\char`\'}} to translate and carry
the default Isar commands \emph{inside} the \isa{\isacommand{C}{\isacartoucheopen}\ {\isachardot}{\isachardot}\ {\isacartoucheclose}} scope, such as \isa{\isacommand{lemma}} or
\isa{\isacommand{by}}. Using \isa{\isacommand{{\isasymsimeq}setup}}, one can even
define an annotation command \isa{\isacommand{C}} taking a C code as argument, as the ML
code of \isa{\isacommand{{\isasymsimeq}setup}} has type
$\alpha^{\text{AST}}~\isatt{{\char`\-}{\char`\>}\ env\ {\char`\-}{\char`\>}}~R_\text{\emph{cmd}}$ (which is enough for
calling \isatt{C{\char`\_}Annotation.command{\char`\'}} in the ML code). Here, whereas the type
\isatt{env} is always the same, the type $\alpha^{\text{AST}} \subseteq
\text{AST}^\text{C11}$ varies depending on {\footnotesize\texttt{<navigation-expr>}} (see \csname isaDof.ref\endcsname[type={text}]{annot2}).%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkuptext}%
Note, however, that the user experience of the IDE changes when nesting commands too
deeply. In terms of error handling and failure treatment, there are some noteworthy implementation
differences between the outermost commands and C annotation commands. Naturally, the PIDE toplevel
has been optimized to maximize the error recovery and parallel execution. Inside a command, the
possibilities to mimic this behaviour are somewhat limited. Whereas the failure of an outermost
command must always be treated as an anomaly, in $\mathcal{L}_\text{annot}$, we are supposed to be
editing in a comment space: an error should be reported but not too much interruptive, because a
comment can contain for good reasons human narrative sentences. On the other hand, an error masking
itself shyly as warning can be overlooked easily, so a form of per-case compromise has to be found
to control the locally permissive set of errors. As a workaround useful during development and
debugging, we offer a further pragma for a global annotation, namely \isa{{\isacharasterisk}}
(in complement to the violet \isa{{\isacharat}}), that controls a switch between a
strict and a permissive error handling for nested annotation commands:
\begin{isar}
    <annotation-comment> $=$ // $[$ * $|$ @ $]^*$ $\mathcal{L}_\text{annot}$
    <annotation-comment> $=$ /* $[$ * $|$ @ $]^*$ $\mathcal{L}_\text{annot}$ */
\end{isar}%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkupsubsection*}%
[label = {annot2},type = {scholarly_paper.technical}, args={label = {annot2},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Evaluation Order%
\end{isamarkupsubsection*}\isamarkuptrue%
%
\begin{isamarkupfigure*}%
[label = {C-sample5},type = {Isa_COL.figure}, args={label = {C-sample5},type = {Isa_COL.figure}, Isa_COL.figure.relative_width = {62}, Isa_COL.figure.src = {figures/A-C-Source5}, Isa_COL.figure.spawn_columns = {True}}]Creating and nesting C annotations: the importance of evaluation order in annotations%
\end{isamarkupfigure*}\isamarkuptrue%
%
\begin{isamarkuptext}%
We will now explain why positional languages are affecting the evaluation time of
annotation commands in \csname isaDof.ref\endcsname[type={Isa_COL.figure}]{C-sample3}. This requires a little zoom on how the
parsing is actually executed.

@@

Note that the C11 grammar has enough scoping structure for the full inference of the C environment
\isatt{env} be at bottom-up time. In terms of efficiency, we use specific
\emph{static} rule wrappers having the potential of overloading default grammar rules (see \csname isaDof.ref\endcsname[type={Isa_COL.figure}]{architecture}), to assign a wrapper to be always executed as soon as a Shift-Reduce
rule node of interest is encountered. The advantage of this construction is that the wrappers are
statically compiled, which results in a very efficient reporting of C type information.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkupsection*}%
[label = {backends},type = {scholarly_paper.technical}, args={label = {backends},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Semantic Back-Ends%
\end{isamarkupsection*}\isamarkuptrue%
%
\begin{isamarkuptext}%
In this section, we briefly present two integrations of verification back-ends for C. We
chose Clean \cite{journals/afp/TuongW19}~\footnote{Stemming from recent HOL-TestGen distributions:
    \url{https://www.brucker.ch/projects/hol-testgen/index.en.html}} used
for program-based test generation \cite{DBLP:conf/tap/Keller18}, and AutoCorres \cite{DBLP:conf/pldi/GreenawayLAK14}, arguably the most developed deductive verification environment
for machine-oriented C available at present.

Note that we were focusing on keeping modifications of integrated components minimal, particularly
for the case of AutoCorres. Certain functionalities like position propagation of HOL terms in
annotations, or ``automatic'' incremental declarations~\footnote{\url{https://github.com/seL4/l4v/blob/master/tools/autocorres/tests/examples/Incremental.thy}}
may require internal revisions on the back-end side. This is out of the scope of this paper.%
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkupsubsection*}%
[label = {clean},type = {scholarly_paper.technical}, args={label = {clean},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]A Simple Typed Memory Model: Clean%
\end{isamarkupsubsection*}\isamarkuptrue%
%
\begin{isamarkuptext}%
Clean (pronounced as: ``C lean'' or ``Céline'' [selin]) is based on a simple,
shallow-style execution model for an imperative target language.
Importing Clean into a theory with its activated back-end proceeds as in \autoref{fig:clean}.
For a deep presentation of Clean, the reader is referred to the published AFP document \cite{journals/afp/TuongW19}.

@@

[label = {autocorres},type = {scholarly_paper.technical}, args={label = {autocorres},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]The Case of AutoCorres%

@@

[label = {securify},type = {scholarly_paper.technical}, args={label = {securify},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Yet Another C Back-End: Securify%

@@

In the Securify project at the Nanyang Technological University \cite{DBLP:conf/tacas/SananZHZTL17,DBLP:conf/itp/HouSTL17}, an extension of L4.verified is in

@@

the verification framework. This ongoing project is already profiting from the overall Isabelle/C
framework, including the integrated parser $\text{PARSE}_{\text{\emph{report}}}^{\text{C11}}$. Note,
however, that the XtratuM micro-kernel \cite{DBLP:journals/sigbed/CarrascosaCMBC14} in study is
using certain advance features of C11 not found in seL4 (including nested functions, or particular

@@

[label = {concl},type = {scholarly_paper.conclusion}, args={label = {concl},type = {scholarly_paper.conclusion}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.text_section.main_author = {}}]Conclusions%

@@

Isabelle/C also forms a basis for future semantically well-understood combinations of back-ends 
based on different semantic interpretations: inside Isabelle, bridge lemmas can be derived 
that describe the precise conditions under which results from one back-end can be re-interpreted
and used in another. Future tactic processes based on these bridge lemmas may open up novel ways 
for semantically safe tool combinations.%
\end{isamarkuptext}\isamarkuptrue%
%
\paragraph*{Acknowledgments.}
\begin{isamarkuptext}%
The authors warmly thank David Sanán and Yang Liu for encouraging the development and reuse of
$C^\downarrow$, initially started in the Securify project \cite{DBLP:conf/tacas/SananZHZTL17}
(\url{http://securify.sce.ntu.edu.sg/}).
\end{isamarkuptext}\isamarkuptrue%
%
\begin{isamarkupsection*}%
[label = {appendix},type = {scholarly_paper.technical}, args={label = {appendix},type = {scholarly_paper.technical}, Isa_COL.text_element.level = {}, Isa_COL.text_element.referentiable = {False}, Isa_COL.text_element.variants = {{STR ''outline'', STR ''document''}}, scholarly_paper.text_section.main_author = {}, scholarly_paper.text_section.fixme_list = {}, Isa_COL.text_element.level = {}, scholarly_paper.technical.definition_list = {}}]Appendix%
\end{isamarkupsection*}\isamarkuptrue%
%
\isadelimdocument
%
\endisadelimdocument
%
\isatagdocument
%
\isamarkupsubsection{Portability of Isabelle/C's Generated Files%
}
\isamarkuptrue%
%
\endisatagdocument
{\isafolddocument}%
%
\isadelimdocument
%
\endisadelimdocument
%
\begin{isamarkuptext}%
To evaluate the portability of the generated files in \csname isaDof.ref\endcsname[type={Isa_COL.figure}]{architecture}, we recall that:

@@

  two ASTs was merged together while releasing version 1.18.0.} Consequently, upgrading
Haskabelle to use the new 1.18.0 represents an entire challenge. Hopefully, it is a task we already
completed while integrating Haskabelle's output with Citadelle's input. Otherwise, we have not
encountered any problems in upgrading Haskabelle from 1.18.0 to more recent versions (such as 1.20.1
which was straightforward).%

@@

evaluation side (with the option \isa{quick{\isacharminus}and{\isacharminus}dirty} activated). In a nutshell, our new
implemented \isa{\isacommand{Haskell{\isacharunderscore}file}} command accepts other various
options:~\footnote{Note that \isa{\isacommand{old{\isacharunderscore}datatype}} is a command,
  whereas \isa{\isakeyword{datatype{\isacharunderscore}old}} is an option provided to the command

@@

The advantage of this construction is the possibility to do immediate code value transformations
obtained from Haskabelle: operationally, \isa{\isacommand{Haskell{\isacharunderscore}file}} makes an
